DP problems of interest

### 2012-09-28 docs/roadmap/problems.tex
Nits.
ideas and problems
more problems, more ideas
Some specs for problems.

%\oddsidemargin   0.0cm
%\evensidemargin  0.0cm
%\topmargin       0.0cm
%\headheight      0.0cm
%\headsep         1.0cm
%\textheight     21.0cm
%\textwidth      16.0cm
\parskip         0.1cm
\parindent       0.0cm
%\footskip        1.0cm

@@

\section{Introduction}
\subsection{Smith-Waterman}
Problem: matching two strings $S$, $T$.

@@

\newpage
\subsection{Nussinov algorithm}
Problem: folding a RNA string $S$ over itself.

@@

% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
%\newpage
%\subsection{Zuker folding}
%Problem: folding a RNA string $S$ over itself.
%
%Alphabets:\ul
%\item Input: $\Sigma(S)=\{A,C,G,U\}$.
%\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
%\item Backtrack matrix: $\Sigma(B)=\{\}$
%\ule
%
%Initialization:\ul
%\item Cost matrices:\ul
%	\item $\Sigma(W) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
%	\item x
%	\item x
%\ule
%\item Backtrack matrix
%\ule
%
%Recurrence:
%\[\begin{array}{rcl}
%W_{(i,j)}&=&\min\left\{\begin{array}{l}
%	W_{(i+1,j)}+b \\
%	W_{(i,j-1)}+b \\
%	V_{(i,j)}+\delta(S_i,S_j) \\
%	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
%\end{array}\right.\\\\
%V_{(i,j)}&=&\min\left\{\begin{array}{ll}
%	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
%	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
%	V_{(i+1,j-1)}+es(i,j) \\
%	VBI(i,j) \\
%	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
%\end{array}\right.\\\\
% F_{(j)}&=&\min\left\{\begin{array}{ll}
%F_{(j-i)} \\
%\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
%\end{array}\right. \text{(Free Energy)}\\\\
%VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
%\end{array}\]
%
%XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...

>>>>

\oddsidemargin   0.0cm
\evensidemargin  0.0cm
\topmargin       0.0cm
\headheight      0.0cm
\headsep         1.0cm
\textheight     21.0cm
\textwidth      16.0cm
\parskip         0.1cm
\parindent       0.0cm
\footskip        1.0cm

@@

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Introduction}
\subsection{Definitions}\ul
\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
\ule

\subsection{Problems classification}
In the literature dynamic programming problems (DP) are classified according to two criteria:

\textbf{Monadic / polyadic}\ul
\item Monadic: on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
\ule

\textbf{Serial/non-serial} \ul
\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial.
\item \textbf{Non-serial:} when the solution of a level depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
\ule

\subsection{Simplifications}
In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example, we can transform
	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
Where $C$ is a matrix that stores the maximum along the column and $A$ is a matrix that stores the sum of the array of the previous elements. They can be easily computed with the additional recurrence.
	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]

This simplification avoids the non-serial dependencies at the cost of an additional storage at the wavefront.

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Problems of interest}
We describe problems structures: inputs, cost matrices, backtracking matrix. They all have an alphabet (that hopefully corresponds to a maximal bit-size) and matrixes dimensions are specified by number of indices and their number of elements is usually the same as the input. Unless otherwise specified, number are all unsigned integers. We describe the problem processing in terms of both initialization and recurrences.

% ----------------------------------------------
\subsection{Smith-Waterman}
Problem: matching two strings $S$, $T$.

@@

% ----------------------------------------------
\subsection{Smith-Waterman with gap extension at different cost}
Problem: matching two strings $S$, $T$.

Initialization:\ul
\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
\item Gap opening matrix: $E_{(i,0)}= 0, 0 \le i \le {\rm length}(S)$
\item Gap extending matrix: $E_{(0,j)}= 0, 0 \le j \le {\rm length}(T)$
\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
\ule

Recurrence for the cost matrix:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)} \]

Recurrence for the gap opening/extending matrices:
\[E_{(i,j)}=\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)} \]

\[F_{(i,j)}=\max\left\{\begin{array}{l|l}
	M_{(i-1, j)} - \alpha & NW\\
	E_{(i-1,j)} - \beta & N\\
\end{array}\right\}=B_{(i,j)} \]

Otherwise written as:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	\max_{1 \le k \le j-1} M_{(i,k)} - \alpha - (j-1-k) \cdot \beta & N\\
	\max_{1 \le k \le i-1} M_{(k,j)} - \alpha - (i-1-k) \cdot \beta & W\\
\end{array}\right\}=B_{(i,j)} \]

% ----------------------------------------------
\newpage
\subsection{Nussinov algorithm}
Problem: folding a RNA string $S$ over itself.

@@

% ------------------------------------------------------------------------------------------------
%\newpage
%\subsection{Zuker folding}
% XXX: we let this aside for the moment as the recurrence has too many unknown parameters
%Problem: folding a RNA string $S$ over itself.
%
%Alphabets:\ul
%\item Input: $\Sigma(S)=\{A,C,G,U\}$.
%\item Cost matrix: $\Sigma(W)=\Sigma(V)=\Sigma(F)=\{0..n\},n=$
%\item Backtrack matrix: $\Sigma(B)=\{\}$
%\ule
%
%Initialization:\ul
%\item Cost matrices:\ul
%	\item xxx
%\ule
%\item Backtrack matrix: $B$
%\ule
%
%Recurrence:
%\[\begin{array}{rcl}
%W_{(i,j)}&=&\min\left\{\begin{array}{l}
%	W_{(i+1,j)}+b \\
%	W_{(i,j-1)}+b \\
%	V_{(i,j)}+\delta(S_i,S_j) \\
%	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)}
%\end{array}\right.\\\\
%V_{(i,j)}&=&\min\left\{\begin{array}{ll}
%	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair}\\
%	eh(i,j)+b \qquad\qquad\, \text{otherwise} \\
%	V_{(i+1,j-1)}+es(i,j) \\
%	VBI(i,j) \\
%	\min_{i<k<j-1}\{W_{(i+,k)}+W_{(k+1,j-1)}\} +c
%\end{array}\right.\\\\
% F_{(j)}&=&\min\left\{\begin{array}{ll}
%F_{(j-i)} \\
%\min_{1\le i< j} (V_{(i,j)} + F_{(j-1)})
%\end{array}\right. \text{(Free Energy)}\\\\
%VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c
%\end{array}\]
%
%XXX: missing: eh(i,j), ebi(i,j,i',j'), es(i,j), ...
